---
title: "A Taxonomy of Failure Modes in a Sub-3B Parameter Agentic AI System on Edge Hardware"
author: "Martin Cote -- Independent Researcher (martin.cote@gmail.com)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Agentic AI systems that orchestrate tool-calling loops via large language models
  (LLMs) exhibit failure patterns that differ substantially from single-turn
  conversational systems. I present an empirically derived taxonomy of failure
  modes observed in bouc.io, a self-hosted agentic platform running Qwen3.5:2b
  (a sub-3B parameter open-weight LLM) on a resource-constrained NVIDIA Jetson
  Orin Nano Super edge device. Failures are classified across six top-level
  categories, Planning, Tool Execution, Model Output Quality, Context/Memory,
  Infrastructure, and User Expectation, each with fine-grained subcategories.
  The taxonomy is derived from 200 runs collected via semi-naturalistic opportunistic
  sampling combining purposive test queries and organic platform use sessions,
  rated by the system operator using a thumbs-up/down feedback mechanism with
  free-text comments. I analyze failure frequency, co-occurrence with tool usage
  patterns, and correlations with run duration and token consumption. The findings
  highlight that planning failures, particularly looping and wrong-tool selection,
  account for the majority (47\%) of coded failures, that negative-feedback runs consume
  45.8\% more tokens than successful ones, and that no technical infrastructure
  failures were observed on the Jetson hardware across this dataset.
output:
  pdf_document:
    latex_engine: pdflatex
    number_sections: true
    fig_caption: true
    toc: false
    keep_tex: false
geometry: "margin=0.7in, top=0.75in, bottom=0.75in"
fontsize: 10pt
classoption: twocolumn
bibliography: references.bib
csl: ieee.csl
link-citations: true
header-includes:
  - \usepackage{setspace}
  - \setstretch{1.0}
  - \usepackage{titlesec}
  - \titlespacing*{\section}{0pt}{5pt}{2pt}
  - \titlespacing*{\subsection}{0pt}{3pt}{1pt}
  - \titleformat{\section}{\normalfont\bfseries}{\thesection.}{0.4em}{}
  - \titleformat{\subsection}{\normalfont\itshape\bfseries}{\thesubsection}{0.4em}{}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{\small\textit{Agent Failure Taxonomy}}
  - \fancyhead[R]{\small M. Cote, 2026}
  - \fancyfoot[C]{\small\thepage}
  - \renewcommand{\headrulewidth}{0.4pt}
  - \usepackage{microtype}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \setlength{\parskip}{3pt}
  - \setlength{\parindent}{0pt}
  - \setlength{\columnsep}{0.2in}
  - \AtBeginDocument{\small}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  fig.width  = 3.2,
  fig.height = 2.4
)

# ── Package loading ──────────────────────────────────────────────────────────
required <- c("dplyr", "ggplot2", "tidyr", "stringr", "knitr",
              "scales", "forcats", "kableExtra")
missing  <- required[!sapply(required, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing, repos = "https://cloud.r-project.org")

library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(knitr)
library(scales)
library(forcats)
library(kableExtra)

# ── Theme ─────────────────────────────────────────────────────────────────────
theme_paper <- function() {
  theme_minimal(base_size = 7) +
    theme(
      panel.grid.minor  = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title        = element_text(size = 6),
      plot.title        = element_text(size = 7, face = "bold"),
      legend.text       = element_text(size = 6),
      legend.key.size   = unit(0.3, "cm")
    )
}
```

```{r load-data}
# ── Load data ─────────────────────────────────────────────────────────────────
# Loads ALL runs_raw_YYYY-MM-DD_HH-MM-SS.csv files from data/ and combines them.
# Each file is the incremental batch produced by collect_metrics.sh for that run.
# Failure codes (failure_category, failure_subcategory, coder_notes) are filled
# in-place in each batch file after collection — no separate coded CSV needed.
# Falls back to the template CSV so the document still compiles with no data.

batch_files <- list.files(
  "data",
  pattern    = "^runs_raw_\\d{4}-\\d{2}-\\d{2}_\\d{2}-\\d{2}-\\d{2}\\.csv$",
  full.names = TRUE
)

read_batch <- function(f) {
  df <- read.csv(f, stringsAsFactors = FALSE, na.strings = c("", "NA"))
  if (nrow(df) == 0) return(NULL)   # skip header-only files
  df
}

non_empty <- Filter(Negate(is.null), lapply(batch_files, read_batch))

if (length(non_empty) > 0) {
  runs <- dplyr::bind_rows(non_empty)
} else {
  # No data yet — fall back to template so the document still compiles
  runs <- read.csv("data/runs_template.csv", stringsAsFactors = FALSE,
                   na.strings = c("", "NA"))
}

# Ensure expected columns exist (tolerates template with zero rows)
expected_cols <- c(
  "run_id", "status", "duration_ms",
  "prompt_tokens", "completion_tokens", "total_tokens", "llm_calls",
  "plan_step_count", "tool_call_count", "failed_tool_count",
  "feedback_rating", "feedback_comment", "model_name",
  "failure_category", "failure_subcategory"
)
for (col in expected_cols) {
  if (!col %in% names(runs)) runs[[col]] <- NA
}

# Derived columns
runs <- runs %>%
  mutate(
    duration_s        = as.numeric(duration_ms) / 1000,
    prompt_tokens     = as.numeric(prompt_tokens),
    completion_tokens = as.numeric(completion_tokens),
    total_tokens      = as.numeric(total_tokens),
    llm_calls         = as.numeric(llm_calls),
    token_efficiency  = completion_tokens / pmax(total_tokens, 1),
    has_feedback    = !is.na(feedback_rating),
    is_negative     = feedback_rating == "negative",
    is_positive     = feedback_rating == "positive",
    run_failed      = status == "failed",
    is_problematic  = (is_negative | run_failed) & !is.na(status),
    failure_cat_lbl = case_when(
      failure_category == "P" ~ "Planning",
      failure_category == "T" ~ "Tool Exec.",
      failure_category == "M" ~ "Model Quality",
      failure_category == "C" ~ "Context/Memory",
      failure_category == "I" ~ "Infrastructure",
      failure_category == "U" ~ "User Expectation",
      TRUE                    ~ NA_character_
    )
  )

# Summary counts
n_total      <- nrow(runs)
n_rated      <- sum(runs$has_feedback, na.rm = TRUE)
n_positive   <- sum(runs$is_positive, na.rm = TRUE)
n_negative   <- sum(runs$is_negative, na.rm = TRUE)
n_failed     <- sum(runs$run_failed, na.rm = TRUE)
n_coded      <- sum(!is.na(runs$failure_category))
```

# Introduction

Large language model (LLM) agents that plan and invoke external tools introduce
failure modes that go well beyond the semantic errors of single-turn chat systems.
Prior work has catalogued failures in cloud-hosted frontier models [@react; @toolformer],
but the behavior of *sub-3B* parameter open-weight models in agentic settings on
resource-constrained hardware is poorly understood as far as I can tell. Small models have been shown
to achieve only 20–40% accuracy on complex agentic tasks [@sub3b], yet the
specific failure patterns, and their operational triggers, have not been
systematically characterized in a production deployment.

I address this gap with a taxonomy derived from semi-naturalistic opportunistic
sampling of bouc.io [@boucio], a self-hosted agentic platform running Qwen3.5:2b
[@qwen2] on an NVIDIA Jetson Orin Nano Super [@jetson]. Rather than constructing
an artificial benchmark, I exploit in-situ user feedback (thumbs-up/down with
free-text comment) collected over `r n_total` runs spanning `r n_rated` rated
interactions.

# Related Work

ReAct [@react] introduced the interleaved reason-act paradigm that underlies
most current tool-calling agents. ToolFormer [@toolformer] showed that LLMs can
learn to call APIs self-supervised, but analysis focused on correctness rather
than failure characterization. AgentBench [@agentbench] benchmarks agent
performance across tasks, noting that smaller models suffer disproportionate
planning failures. HuggingGPT [@hugginggpt] and similar orchestration frameworks
identify tool-selection errors as a dominant failure class. ToolLLM [@toolllm]
benchmarks tool-calling across 16,000+ real-world APIs, finding that even large
models struggle with multi-hop chains — a pattern my data confirms at sub-3B
scale. Sumers et al. [@cogarch] provide a comprehensive taxonomy of cognitive
architectures for language agents, within which the planner-executor pattern of
bouc.io corresponds to a plan-then-act decomposition. MemGPT [@memgpt] addresses
context limitations via hierarchical memory management; the failure modes it
targets (C-MM, C-OV) remain latent in our deployment due to current memory system
constraints (see Discussion). My work differs in three ways: (1) it uses a sub-3B
model, (2) it is grounded in a production-like deployment on edge GPU hardware, and
(3) failure labels come from the operator rather than an automated evaluator.

# Methodology

## System Overview

bouc.io implements a two-phase planner-executor agent [@boucio]. The *planning
phase* prompts the LLM to produce a structured JSON plan (goal + ordered steps
with tool intents). The *execution phase* iterates over steps, invoking tools and
feeding results back into context. Fourteen tools are available: `web_search`,
`calculator`, `http_request`, `echo`, `fetch_page`, `json_query`, `text_extract`, `memory_search`, `bash`, `read_file`, `write_file`, `glob_files`, `list_dir`, and `time_date`. Each run is stored in
PostgreSQL with full provenance: plan JSON, per-step LLM inputs/outputs, tool
call inputs/outputs, token counts, and timing.

## Observed Performance
\vspace{-8pt}
```{r perf-table}
# Use the latest alias written by collect_metrics.sh; fall back to most recent
# dated snapshot if the alias doesn't exist yet.
perf_csv <- "data/performance_metrics.csv"
if (!file.exists(perf_csv)) {
  dated <- sort(list.files("data", pattern = "^performance_metrics_.*\\.csv$",
                            full.names = TRUE), decreasing = TRUE)
  if (length(dated) > 0) perf_csv <- dated[1]
}
if (file.exists(perf_csv)) {
  perf <- read.csv(perf_csv, stringsAsFactors = FALSE)
  # Show service label only on first row of each group
  perf$service[duplicated(perf$service)] <- ""
  knitr::kable(
    perf[, c("service", "metric", "value", "unit")],
    col.names = c("Service", "Metric", "Value", "Unit"),
    booktabs  = TRUE,
    format    = "latex",
    caption   = "Observed operational performance measurements.",
    align     = c("l", "l", "r", "l"),
    linesep   = ""
  ) %>%
    kable_styling(
      latex_options = c("hold_position", "scale_down"),
      full_width    = FALSE
    )
}
```
† Chatbot latency measured with thinking enabled; p95 driven by reasoning chain length variance, not I/O or retrieval overhead.

## Data Collection

After each agent response, the operator rated the run with a thumbs-up (positive)
or thumbs-down (negative) button and optionally provided a free-text comment
explaining the failure. This feedback is stored in the `RunFeedback` table linked
one-to-one to each `Run`. Data was extracted incrementally via the SQL query in
`data/extract_runs.sql`, joining `Run`, `Assignment`, `Message`, `ToolCall`,
`RunStep`, and `RunFeedback`. Each extraction batch is saved as a dated CSV
(`runs_raw_YYYY-MM-DD_HH-MM-SS.csv`); the Rmd loads and combines all batches
at render time. Queries were a mix of deliberate tool-exercising prompts (see
Appendix) and sessions arising from routine platform use.

The resulting dataset contains **`r n_total` runs** of which **`r n_rated`** are
rated (**`r n_positive` positive**, **`r n_negative` negative**).
An additional **`r n_failed`** runs ended in technical failure (`status = 'failed'`).

## Coding Procedure

Two sources were coded: (1) all runs with `feedback_rating = 'negative'` and
(2) all runs with `status = 'failed'` regardless of feedback. Coding was
performed by the operator in-place within each dated batch CSV after extraction.
Each run received one *top-level category* (P, T, M, C, I, or U) and one
*subcategory* (e.g., P-LP for planning loop). In cases of multiple failures,
the primary cause was coded. **`r n_coded` runs** received failure codes,
corresponding to all `r n_negative` negative-feedback runs in this dataset
(no technical failures were observed).

# Taxonomy

Table 1 presents the full taxonomy. Six top-level categories were identified,
with 20 subcategories total.

```{r taxonomy-table, results='asis'}
taxonomy <- data.frame(
  Code = c(
    "P", "P-NJ", "P-OD", "P-WT", "P-LP",
    "T", "T-NF", "T-TO", "T-PE", "T-EX", "T-NE",
    "M", "M-HL", "M-RP", "M-INC", "M-RF",
    "C", "C-MM", "C-OV",
    "I", "I-OOM", "I-NET", "I-DB",
    "U", "U-MI", "U-IN"
  ),
  Description = c(
    "Planning failures", "No valid JSON plan produced",
    "Over-decomposition (too many steps)", "Wrong tool selected in plan",
    "Planning loop / duplicate plan",
    "Tool execution failures", "Tool not found (registry mismatch)",
    "Tool execution timeout", "Tool output parsing error",
    "Tool raised exception", "Tool not executed (execution system skipped step)",
    "Model output quality", "Hallucinated fact or tool name",
    "Repetitive / looping response", "Incomplete response (cut off)",
    "Model refusal",
    "Context / memory", "Wrong memories injected",
    "Context overflow (prompt truncated)",
    "Infrastructure", "Out-of-memory (Jetson OOM)",
    "Network / Redis error", "Database error",
    "User expectation gap", "Misunderstood intent",
    "Insufficient depth or detail"
  ),
  stringsAsFactors = FALSE
)

# Indent subcategories
taxonomy$Code <- ifelse(
  nchar(taxonomy$Code) > 1,
  paste0("  ", taxonomy$Code),
  taxonomy$Code
)

knitr::kable(
  taxonomy,
  format  = "latex",
  booktabs = TRUE,
  caption = "Failure taxonomy (6 top-level categories, 20 subcategories).",
  col.names = c("Code", "Description"),
  linesep = ""
) %>%
  kable_styling(font_size = 6, latex_options = c("hold_position", "scale_down")) %>%
  row_spec(c(1, 6, 12, 17, 20, 24), bold = TRUE)
```

# Results

Across all completed runs, the median total token consumption was
`r median(runs$total_tokens, na.rm = TRUE)` tokens
(IQR: `r quantile(runs$total_tokens, 0.25, na.rm = TRUE)`--`r quantile(runs$total_tokens, 0.75, na.rm = TRUE)`).
Negative-feedback runs consumed a median of
`r median(runs$total_tokens[runs$is_negative == TRUE], na.rm = TRUE)` tokens versus
`r median(runs$total_tokens[runs$is_positive == TRUE], na.rm = TRUE)` for positive runs,
suggesting that failure-prone interactions involve longer context accumulation across
planning and execution phases.

## Overall Run Quality

Figure 1 shows the distribution of run outcomes across the full dataset of `r n_total` runs. Runs are classified as *positive feedback*, *negative feedback*, *failed* (status = 'failed'), or *no feedback* (unrated).

```{r quality-bar, fig.cap="Run outcome distribution."}
if (n_total > 0) {
  quality_df <- runs %>%
    mutate(outcome = case_when(
      run_failed         ~ "Failed",
      is_negative        ~ "Negative feedback",
      is_positive        ~ "Positive feedback",
      TRUE               ~ "No feedback"
    )) %>%
    count(outcome) %>%
    mutate(pct = n / sum(n),
           outcome = fct_relevel(outcome, "Positive feedback",
                                 "No feedback", "Negative feedback", "Failed"))

  ggplot(quality_df, aes(x = outcome, y = pct, fill = outcome)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = paste0(n, "\n(", percent(pct, 1), ")")),
              vjust = -0.3, size = 2) +
    scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
    scale_fill_manual(values = c(
      "Positive feedback" = "#4caf50",
      "No feedback"       = "#9e9e9e",
      "Negative feedback" = "#ff9800",
      "Failed"            = "#f44336"
    )) +
    labs(x = NULL, y = "Proportion of runs", title = "Run Outcomes") +
    theme_paper() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 6))
} else {
  ggplot() + annotate("text", x=1, y=1, label="No data yet — collect runs first") +
    theme_void()
}
```

## Failure Category Frequency

Figure 2 shows the share of coded failures attributed to each of the six top-level taxonomy categories. Only the `r n_coded` runs that received a failure code contribute to this chart.

```{r category-freq, fig.cap="Frequency of failure categories among coded runs."}
coded_runs <- runs %>% filter(!is.na(failure_category))

if (nrow(coded_runs) > 0) {
  cat_counts <- coded_runs %>%
    count(failure_cat_lbl) %>%
    filter(!is.na(failure_cat_lbl)) %>%
    mutate(pct = n / sum(n),
           failure_cat_lbl = fct_reorder(failure_cat_lbl, n))

  ggplot(cat_counts, aes(x = failure_cat_lbl, y = pct)) +
    geom_col(fill = "#1565c0", width = 0.6) +
    geom_text(aes(label = paste0(n, " (", percent(pct, 1), ")")),
              hjust = -0.1, size = 2) +
    coord_flip() +
    scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
    labs(x = NULL, y = "Share of coded failures",
         title = "Failure Category Distribution") +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No coded failures yet") +
    theme_void()
}
```

## Run Duration by Outcome

Figure 3 compares run duration in seconds across outcome categories. Longer durations in negative or failed runs reflect additional planning and execution cycles before the agent terminates or the user abandons the interaction.

```{r duration-box, fig.cap="Run duration (s) by outcome category. Horizontal bars show medians."}
dur_df <- runs %>%
  filter(!is.na(duration_s)) %>%
  mutate(outcome = case_when(
    run_failed  ~ "Failed",
    is_negative ~ "Negative",
    is_positive ~ "Positive",
    TRUE        ~ "No feedback"
  ))

if (nrow(dur_df) > 0) {
  ggplot(dur_df, aes(x = outcome, y = duration_s, fill = outcome)) +
    geom_boxplot(outlier.size = 0.8, width = 0.5, show.legend = FALSE) +
    scale_fill_manual(values = c(
      "Positive"    = "#4caf50",
      "Negative"    = "#ff9800",
      "Failed"      = "#f44336",
      "No feedback" = "#9e9e9e"
    )) +
    labs(x = NULL, y = "Duration (seconds)", title = "Run Duration by Outcome") +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No duration data yet") +
    theme_void()
}
```

## LLM Call Count Distribution

Figure 4 shows the distribution of LLM calls per run, stacked by outcome. A high LLM call count for negative runs is a proxy signal for planning loops (P-LP), where the planner repeatedly regenerates steps before the executor detects the cycle.

```{r llm-calls-hist, fig.cap="Distribution of LLM calls per run by outcome."}
llm_df <- runs %>%
  filter(!is.na(llm_calls)) %>%
  mutate(outcome = case_when(
    run_failed  ~ "Failed",
    is_negative ~ "Negative",
    is_positive ~ "Positive",
    TRUE        ~ "No feedback"
  ))

if (nrow(llm_df) > 0) {
  ggplot(llm_df, aes(x = llm_calls, fill = outcome)) +
    geom_histogram(binwidth = 1, position = "stack", colour = "white", linewidth = 0.2) +
    scale_fill_manual(values = c(
      "Positive"    = "#4caf50",
      "Negative"    = "#ff9800",
      "Failed"      = "#f44336",
      "No feedback" = "#9e9e9e"
    )) +
    labs(x = "LLM calls per run", y = "Count",
         title = "LLM Call Distribution", fill = NULL) +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No LLM call data yet") +
    theme_void()
}
```

## Token Usage by Outcome

Figures 5 and 6 quantify token consumption by outcome. Negative-feedback and failed runs are expected to consume more tokens due to additional context accumulation across repeated planning and execution cycles.

```{r token-box, fig.cap="Total tokens per run by outcome. Bars show medians."}
tok_df <- runs %>%
  filter(!is.na(total_tokens), total_tokens > 0) %>%
  mutate(outcome = case_when(
    run_failed  ~ "Failed",
    is_negative ~ "Negative",
    is_positive ~ "Positive",
    TRUE        ~ "No feedback"
  ))

if (nrow(tok_df) > 0) {
  ggplot(tok_df, aes(x = outcome, y = total_tokens, fill = outcome)) +
    geom_boxplot(outlier.size = 0.8, width = 0.5, show.legend = FALSE) +
    scale_fill_manual(values = c(
      "Positive"    = "#4caf50",
      "Negative"    = "#ff9800",
      "Failed"      = "#f44336",
      "No feedback" = "#9e9e9e"
    )) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "Total tokens", title = "Token Usage by Outcome") +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No token data yet") + theme_void()
}
```

```{r token-breakdown, fig.cap="Mean prompt vs completion tokens by outcome."}
if (nrow(tok_df) > 0) {
  tok_long <- tok_df %>%
    group_by(outcome) %>%
    summarise(
      Prompt     = mean(prompt_tokens,     na.rm = TRUE),
      Completion = mean(completion_tokens, na.rm = TRUE),
      .groups    = "drop"
    ) %>%
    pivot_longer(c(Prompt, Completion), names_to = "type", values_to = "tokens")

  ggplot(tok_long, aes(x = outcome, y = tokens, fill = type)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = c("Prompt" = "#5c6bc0", "Completion" = "#26a69a")) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "Mean tokens",
         title = "Prompt vs Completion Tokens", fill = NULL) +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No token data yet") + theme_void()
}
```

## Subcategory Breakdown

Table 2 presents failure counts at the subcategory level. This view identifies which specific failure modes within each top-level category are most prevalent, enabling targeted mitigation prioritization.

```{r subcat-table, results='asis'}
if (nrow(coded_runs) > 0) {
  subcat_tbl <- coded_runs %>%
    filter(!is.na(failure_subcategory)) %>%
    count(failure_category, failure_subcategory) %>%
    arrange(failure_category, desc(n)) %>%
    rename(Category = failure_category,
           Subcategory = failure_subcategory,
           Count = n)

  knitr::kable(
    subcat_tbl,
    format   = "latex",
    booktabs = TRUE,
    caption  = "Failure subcategory counts.",
    linesep  = ""
  ) %>%
    kable_styling(font_size = 6, latex_options = c("hold_position"))
} else {
  cat("*No coded failures yet — add failure codes to a runs\\_raw batch file to see this table.*\n")
}
```

## Tool Usage and Failure Co-occurrence

Figure 7 contrasts tool invocation frequency between runs with positive outcomes ("OK") and runs with negative feedback or failures ("Problematic"). Tools that appear disproportionately in problematic runs may indicate model difficulty with specific tool schemas or output formats.

```{r tool-failure, fig.cap="Tool invocation frequency for positive vs negative runs."}
tool_df <- runs %>%
  filter(!is.na(tools_used) & tools_used != "") %>%
  mutate(outcome = if_else(is_negative | run_failed, "Problematic", "OK")) %>%
  mutate(tool_list = str_split(tools_used, "; ")) %>%
  unnest(tool_list) %>%
  filter(!is.na(tool_list) & tool_list != "")

if (nrow(tool_df) > 0) {
  tool_counts <- tool_df %>%
    count(tool_list, outcome) %>%
    group_by(tool_list) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(tool_list = fct_reorder(tool_list, total))

  ggplot(tool_counts, aes(x = tool_list, y = n, fill = outcome)) +
    geom_col(position = "dodge", width = 0.6) +
    coord_flip() +
    scale_fill_manual(values = c("OK" = "#4caf50", "Problematic" = "#f44336")) +
    labs(x = NULL, y = "Invocation count",
         title = "Tool Usage by Run Outcome", fill = NULL) +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No tool usage data yet") +
    theme_void()
}
```

## Failures by Model

Figures 8 and 9 present outcome and failure-category distributions broken down by model. All runs in the current dataset used `qwen3.5:2b`; cross-model comparison will populate automatically as additional model deployments are instrumented.

```{r model-outcome, fig.cap="Run outcome distribution by model."}
model_df <- runs %>%
  filter(!is.na(model_name) & model_name != "") %>%
  mutate(outcome = case_when(
    run_failed  ~ "Failed",
    is_negative ~ "Negative feedback",
    is_positive ~ "Positive feedback",
    TRUE        ~ "No feedback"
  )) %>%
  count(model_name, outcome) %>%
  group_by(model_name) %>%
  mutate(pct = n / sum(n),
         outcome = fct_relevel(outcome, "Positive feedback",
                               "No feedback", "Negative feedback", "Failed"))

if (nrow(model_df) > 0) {
  ggplot(model_df, aes(x = model_name, y = pct, fill = outcome)) +
    geom_col(position = "stack", width = 0.6) +
    geom_text(aes(label = ifelse(pct > 0.05, paste0(n), "")),
              position = position_stack(vjust = 0.5), size = 2) +
    scale_y_continuous(labels = percent_format()) +
    scale_fill_manual(values = c(
      "Positive feedback" = "#4caf50",
      "No feedback"       = "#9e9e9e",
      "Negative feedback" = "#ff9800",
      "Failed"            = "#f44336"
    )) +
    labs(x = NULL, y = "Proportion of runs",
         title = "Run Outcomes by Model", fill = NULL) +
    theme_paper() +
    theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 6))
} else {
  ggplot() + annotate("text", x=1, y=1, label="No model data yet") + theme_void()
}
```

```{r model-failure-cat, fig.cap="Failure category distribution by model (coded runs only)."}
coded_model_df <- runs %>%
  filter(!is.na(failure_category), !is.na(model_name), model_name != "") %>%
  count(model_name, failure_cat_lbl) %>%
  filter(!is.na(failure_cat_lbl)) %>%
  group_by(model_name) %>%
  mutate(pct = n / sum(n))

if (nrow(coded_model_df) > 0) {
  ggplot(coded_model_df, aes(x = model_name, y = pct, fill = failure_cat_lbl)) +
    geom_col(position = "stack", width = 0.6) +
    geom_text(aes(label = ifelse(pct > 0.05, paste0(n), "")),
              position = position_stack(vjust = 0.5), size = 2) +
    scale_y_continuous(labels = percent_format()) +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = "Share of coded failures",
         title = "Failure Categories by Model", fill = NULL) +
    theme_paper() +
    theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 6))
} else {
  ggplot() + annotate("text", x=1, y=1, label="No coded failures yet") + theme_void()
}
```

# Discussion

```{r discussion-stats, include=FALSE}
# Pre-compute stats for inline use
pct_neg <- if (n_rated > 0) round(n_negative / n_rated * 100, 1) else NA
pct_fail <- if (n_total > 0) round(n_failed / n_total * 100, 1) else NA

top_cat <- coded_runs %>%
  count(failure_cat_lbl) %>%
  arrange(desc(n)) %>%
  slice(1) %>%
  pull(failure_cat_lbl)

top_cat <- if (length(top_cat) == 0) "N/A" else top_cat

# Mann-Whitney U test: token consumption, negative vs positive runs
rated_runs <- runs %>%
  filter(feedback_rating %in% c("positive", "negative"), total_tokens > 0)
if (nrow(rated_runs) > 2) {
  wt     <- wilcox.test(total_tokens ~ feedback_rating, data = rated_runs)
  wt_W   <- round(wt$statistic)
  wt_p   <- format.pval(wt$p.value, digits = 2, eps = 0.001)
} else {
  wt_W <- NA; wt_p <- "N/A"
}
neg_med_tok <- median(runs$total_tokens[runs$is_negative == TRUE], na.rm = TRUE)
pos_med_tok <- median(runs$total_tokens[runs$is_positive == TRUE], na.rm = TRUE)
pct_tok_increase <- if (!is.na(neg_med_tok) && !is.na(pos_med_tok) && pos_med_tok > 0)
  round((neg_med_tok - pos_med_tok) / pos_med_tok * 100) else NA
```

Of the `r n_rated` rated runs, `r pct_neg`% received negative feedback and
`r pct_fail`% ended in technical failure. The most frequent failure category
was **`r top_cat`** (`r if(nrow(coded_runs)>0) round(sum(coded_runs$failure_category=="P",na.rm=TRUE)/nrow(coded_runs)*100,0) else 0`%
of coded failures). These results are consistent with prior reports that
sub-3B models struggle with structured output generation and multi-step
reasoning [@sub3b; @agentbench].

Several patterns warrant highlighting. First, planning loops (P-LP) are the
single dominant failure subcategory, occurring when the planner re-generates
the same step or emits duplicate tool intents in successive LLM calls. This
drives disproportionately high token consumption: negative-feedback runs consumed
a median of `r neg_med_tok` tokens versus `r pos_med_tok` for positive runs —
a **`r pct_tok_increase`% increase** (Mann-Whitney U, W\,=\,`r wt_W`,
p\,`r wt_p`) — as the loop accumulates context before the executor detects the
cycle. Task complexity is a likely confound: more demanding queries may drive
both higher token use and greater failure likelihood. Second, tool output parsing errors (T-PE)
represent the dominant tool-execution failure, arising when the model returns
malformed JSON in a tool call argument rather than within the structured plan.
This is consistent with small-model limitations in maintaining strict output
schemas under long context [@sub3b]. Third, wrong-tool selection (P-WT) occurs
when the planner assigns a `web_search` intent to a query that the `calculator`, `http_request`, `web_search`
or `time_date` tool could resolve deterministically, increasing latency and
introducing retrieval noise.

Notably, no infrastructure failures (I-OOM, I-NET) were observed across `r n_total`
runs, suggesting that the hardware-level stability of the Jetson Orin Nano Super
is adequate for the observed query load, and that the llama3.2-to-qwen3.5:2b
model migration successfully resolved the out-of-memory crashes documented in
earlier operations.  It is important to note, the tests were done under a single user workload, allowing plenty of capacity to process each request in sequence.

The absence of Context/Memory (C) failures warrants explanation. The platform's
memory distiller uses the same Qwen3.5:2b model to evaluate and extract memories,
applying a confidence threshold of 0.5. In practice this produces few high-quality
extractions per run, resulting in sparse memory injection across sessions. With
minimal memories in context, retrieval failures (C-MM) and context overflow (C-OV)
are rarely triggered under current operating conditions. This is a system maturity
limitation rather than evidence that C-category failures are absent in principle, as a more capable memory pipeline would be expected to surface this category
[@memgpt].

The in-situ feedback mechanism proved effective for semi-naturalistic data
collection:
the operator naturally left short comments that described not just *what* failed but
*why*, providing richer signal than a binary label alone. This suggests that
embedding lightweight feedback collection directly in operational AI systems is
a viable alternative to purpose-built evaluation harnesses for longitudinal
failure characterization.

Regarding the model-level analysis: all `r n_total` runs in this dataset used
a single model (`qwen3.5:2b`), so cross-model comparison charts reflect a
single stratum. Comparative analysis across models is reserved for future work
as additional model deployments are instrumented.

## Limitations

This study has several limitations. First, all failure codes were assigned by a
single coder (the system operator, i.e. myself), introducing potential for confirmation bias.
No inter-rater reliability measure (e.g., Cohen's kappa) was computed; internal
consistency was partially mitigated by applying the codebook in a single
continuous session with explicit decision rules for ambiguous cases. Second, the
dataset covers a single model (Qwen3.5:2b) on a single deployment; findings may
not generalise to other sub-3B models or hardware configurations. Third, data
collection blends purposive test queries with organic use sessions, limiting
claims about the representativeness of failure rates relative to unconstrained
deployment.

# Conclusion

I have presented a six-category, 20-subcategory taxonomy of failure modes for
a sub-3B agentic AI system on edge GPU hardware, grounded in `r n_total`
semi-naturalistic runs on a single-node Jetson Orin Nano Super deployment. Planning
failures, particularly looping (P-LP) and wrong-tool selection (P-WT),
account for 47% of coded failures and are the primary driver of elevated token
consumption in problematic runs. Tool output parsing errors (T-PE) represent
the dominant execution-layer failure, consistent with small-model limitations
in strict schema adherence. Infrastructure failures were absent in this dataset,
validating the hardware and model selection decisions made during platform
iteration. The taxonomy provides a structured vocabulary for discussing,
prioritizing, and mitigating agentic system failures in resource-constrained
deployments. The dataset is incremental and will grow as the platform is used;
the analysis pipeline is designed to re-render automatically as new rated runs
are collected and coded. All data, extraction queries, and analysis code are
available in the companion repository.

# Appendix: Test Query Corpus {-}

The following queries were submitted to the agent during data collection. Queries
were chosen to cover the main tools available
(`web_search`, `calculator`, `time_date`, `echo`, `http_request`) as well as
multi-step chaining, ambiguous intent, and out-of-scope requests likely to
surface model quality and user expectation failures.

**Single-tool — web search**

- Search for the weather in Paris.
- Who is the current Prime Minister of Canada?
- What is the capital city of Australia?
- What is the population of Tokyo?
- What is the current stock price of Apple?

**Single-tool — calculator**

- What is 2+2, then multiply that by 3, then add 10 to it.
- What is 15% of 847?
- What is the square root of 144, divided by 3?
- If I invest \$1,000 at 5% annual interest for 10 years, what is the final amount?

**Single-tool — time / date**

- What time is it in Tokyo?
- What time is it right now in London?
- What day of the week will January 1st, 2027 fall on?
- How many days until Christmas?

**Multi-step / tool chaining**

- How long is the drive between Montreal, QC and New York City, NY?
- What is today's date? Then tell me what date it will be in 90 days.
- Search for the speed of light in km/s, then calculate how far light travels in 10 minutes.

**Ambiguous / edge cases**

- Tell me a fun fact.
- What is the meaning of life?
- Summarize the history of the internet in 3 bullet points.

**Out-of-scope / user expectation**

- Write me a Python function to sort a list.
- Translate "hello, how are you?" into French, Spanish, and Japanese.

---

\small
*Platform source code and infrastructure modules for bouc.io are maintained in a series of private
GitLab repositories. Blog posts documenting each major system milestone and specific technical components are
published at \url{https://www.martincote.ca/blogs.html}.*

---

\small
*\textbf{Data \& Code Availability.} The R Markdown source, performance metrics CSV, and
collection scripts for this paper are openly available at
\url{https://github.com/MartinCote1978/agent-failure-taxonomy-research}*

---

\small
*\textbf{AI Usage Disclosure.} This document was created with assistance from AI tools. The content has been reviewed and edited by a human. For more information on the extent and nature of AI usage, please contact the author.*

---


# References {-}
