---
title: "A Taxonomy of Failure Modes in a Sub-3B Parameter Agentic AI System on Edge Hardware"
author: "Martin Cote -- Independent Researcher (martin.cote@gmail.com)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
  Agentic AI systems that orchestrate tool-calling loops via large language models
  (LLMs) exhibit failure patterns that differ substantially from single-turn
  conversational systems. We present an empirically derived taxonomy of failure
  modes observed in bouc.io, a self-hosted agentic platform running Qwen3.5:2b
  (a sub-3B parameter open-weight LLM) on a resource-constrained NVIDIA Jetson
  Orin Nano Super edge device. Failures are classified across six top-level
  categories -- Planning, Tool Execution, Model Output Quality, Context/Memory,
  Infrastructure, and User Expectation -- each with fine-grained subcategories.
  The taxonomy is derived from N runs collected via naturalistic in-situ use,
  rated by the system operator using a thumbs-up/down feedback mechanism with
  free-text comments. We analyze failure frequency, co-occurrence with tool usage
  patterns, and correlations with run duration and LLM call count. Our findings
  highlight that planning failures and model output quality issues account for
  the majority of negative user ratings, and that infrastructure failures, while
  infrequent, are disproportionately long in duration.
output:
  pdf_document:
    latex_engine: pdflatex
    number_sections: true
    fig_caption: true
    toc: false
    keep_tex: false
geometry: "margin=0.7in, top=0.75in, bottom=0.75in"
fontsize: 10pt
classoption: twocolumn
bibliography: references.bib
csl: ieee.csl
link-citations: true
header-includes:
  - \usepackage{setspace}
  - \setstretch{1.0}
  - \usepackage{titlesec}
  - \titlespacing*{\section}{0pt}{5pt}{2pt}
  - \titlespacing*{\subsection}{0pt}{3pt}{1pt}
  - \titleformat{\section}{\normalfont\bfseries}{\thesection.}{0.4em}{}
  - \titleformat{\subsection}{\normalfont\itshape\bfseries}{\thesubsection}{0.4em}{}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{\small\textit{Agent Failure Taxonomy}}
  - \fancyhead[R]{\small M. Cote, 2026}
  - \fancyfoot[C]{\small\thepage}
  - \renewcommand{\headrulewidth}{0.4pt}
  - \usepackage{microtype}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \setlength{\parskip}{3pt}
  - \setlength{\parindent}{0pt}
  - \setlength{\columnsep}{0.2in}
  - \AtBeginDocument{\small}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  fig.width  = 3.2,
  fig.height = 2.4
)

# ── Package loading ──────────────────────────────────────────────────────────
required <- c("dplyr", "ggplot2", "tidyr", "stringr", "knitr",
              "scales", "forcats", "kableExtra")
missing  <- required[!sapply(required, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing, repos = "https://cloud.r-project.org")

library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(knitr)
library(scales)
library(forcats)
library(kableExtra)

# ── Theme ─────────────────────────────────────────────────────────────────────
theme_paper <- function() {
  theme_minimal(base_size = 7) +
    theme(
      panel.grid.minor  = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title        = element_text(size = 6),
      plot.title        = element_text(size = 7, face = "bold"),
      legend.text       = element_text(size = 6),
      legend.key.size   = unit(0.3, "cm")
    )
}
```

```{r load-data}
# ── Load data ─────────────────────────────────────────────────────────────────
# Expects data/runs_coded.csv  (see data/DATA-COLLECTION.md for how to produce it)
# Falls back to the template CSV so the document still compiles with no data.

csv_path <- "data/runs_coded.csv"
if (!file.exists(csv_path)) csv_path <- "data/runs_raw.csv"
if (!file.exists(csv_path)) csv_path <- "data/runs_template.csv"

runs <- read.csv(csv_path, stringsAsFactors = FALSE, na.strings = c("", "NA"))

# Ensure expected columns exist (tolerates template with zero rows)
expected_cols <- c(
  "run_id", "status", "duration_ms", "llm_calls", "plan_step_count",
  "tool_call_count", "failed_tool_count", "feedback_rating", "feedback_comment",
  "failure_category", "failure_subcategory"
)
for (col in expected_cols) {
  if (!col %in% names(runs)) runs[[col]] <- NA
}

# Derived columns
runs <- runs %>%
  mutate(
    duration_s      = as.numeric(duration_ms) / 1000,
    has_feedback    = !is.na(feedback_rating),
    is_negative     = feedback_rating == "negative",
    is_positive     = feedback_rating == "positive",
    run_failed      = status == "failed",
    is_problematic  = (is_negative | run_failed) & !is.na(status),
    failure_cat_lbl = case_when(
      failure_category == "P" ~ "Planning",
      failure_category == "T" ~ "Tool Exec.",
      failure_category == "M" ~ "Model Quality",
      failure_category == "C" ~ "Context/Memory",
      failure_category == "I" ~ "Infrastructure",
      failure_category == "U" ~ "User Expectation",
      TRUE                    ~ NA_character_
    )
  )

# Summary counts
n_total      <- nrow(runs)
n_rated      <- sum(runs$has_feedback, na.rm = TRUE)
n_positive   <- sum(runs$is_positive, na.rm = TRUE)
n_negative   <- sum(runs$is_negative, na.rm = TRUE)
n_failed     <- sum(runs$run_failed, na.rm = TRUE)
n_coded      <- sum(!is.na(runs$failure_category))
```

# Introduction

Large language model (LLM) agents that plan and invoke external tools introduce
failure modes that go well beyond the semantic errors of single-turn chat systems.
Prior work has catalogued failures in cloud-hosted frontier models [@react; @toolformer],
but the behavior of *sub-3B* parameter open-weight models in agentic settings on
resource-constrained hardware is poorly understood. Small models have been shown
to achieve only 20–40% accuracy on complex agentic tasks [@sub3b], yet the
specific failure patterns—and their operational triggers—have not been
systematically characterized in a production deployment.

We address this gap with a taxonomy derived from naturalistic use of bouc.io
[@boucio], a self-hosted agentic platform running Qwen3.5:2b [@qwen2] on an
NVIDIA Jetson Orin Nano Super [@jetson]. Rather than constructing an artificial
benchmark, we exploit in-situ user feedback (thumbs-up/down with free-text
comment) collected over `r n_total` runs spanning `r n_rated` rated interactions.

# Related Work

ReAct [@react] introduced the interleaved reason-act paradigm that underlies
most current tool-calling agents. ToolFormer [@toolformer] showed that LLMs can
learn to call APIs self-supervised, but analysis focused on correctness rather
than failure characterization. AgentBench [@agentbench] benchmarks agent
performance across tasks, noting that smaller models suffer disproportionate
planning failures. HuggingGPT [@hugginggpt] and similar orchestration frameworks
identify tool-selection errors as a dominant failure class. Our work differs in
three ways: (1) it uses a sub-3B model, (2) it is grounded in a production
deployment on edge GPU hardware, and (3) failure labels come from the operator
rather than an automated evaluator.

# Methodology

## System Overview

bouc.io implements a two-phase planner-executor agent [@boucio]. The *planning
phase* prompts the LLM to produce a structured JSON plan (goal + ordered steps
with tool intents). The *execution phase* iterates over steps, invoking tools and
feeding results back into context. Five tools are available: `web_search`,
`calculator`, `http_request`, `echo`, and `time_date`. Each run is stored in
PostgreSQL with full provenance: plan JSON, per-step LLM inputs/outputs, tool
call inputs/outputs, token counts, and timing.

## Data Collection

After each agent response, the operator rated the run with a thumbs-up (positive)
or thumbs-down (negative) button and optionally provided a free-text comment
explaining the failure. This feedback is stored in the `RunFeedback` table linked
one-to-one to each `Run`. Data was extracted via the SQL query in
`data/extract_runs.sql`, joining `Run`, `Assignment`, `Message`, `ToolCall`,
`RunStep`, and `RunFeedback`.

The resulting dataset contains **`r n_total` runs** of which **`r n_rated`** are
rated (**`r n_positive` positive**, **`r n_negative` negative**).
An additional **`r n_failed`** runs ended in technical failure (`status = 'failed'`).

## Coding Procedure

Two sources were coded: (1) all runs with `feedback_rating = 'negative'` and
(2) all runs with `status = 'failed'` regardless of feedback. Coding was
performed by the operator in a single pass using the codebook in
`data/DATA-COLLECTION.md`. Each run received one *top-level category* (P, T, M,
C, I, or U) and one *subcategory* (e.g., P-OD for over-decomposition). In cases
of multiple failures, the primary cause was coded. **`r n_coded` runs** received
failure codes.

# Taxonomy

Table 1 presents the full taxonomy. Six top-level categories were identified,
with 19 subcategories total.

```{r taxonomy-table, results='asis'}
taxonomy <- data.frame(
  Code = c(
    "P", "P-NJ", "P-OD", "P-WT", "P-LP",
    "T", "T-NF", "T-TO", "T-PE", "T-EX",
    "M", "M-HL", "M-RP", "M-INC", "M-RF",
    "C", "C-MM", "C-OV",
    "I", "I-OOM", "I-NET", "I-DB",
    "U", "U-MI", "U-IN"
  ),
  Description = c(
    "Planning failures", "No valid JSON plan produced",
    "Over-decomposition (too many steps)", "Wrong tool selected in plan",
    "Planning loop / duplicate plan",
    "Tool execution failures", "Tool not found (registry mismatch)",
    "Tool execution timeout", "Tool output parsing error",
    "Tool raised exception",
    "Model output quality", "Hallucinated fact or tool name",
    "Repetitive / looping response", "Incomplete response (cut off)",
    "Model refusal",
    "Context / memory", "Wrong memories injected",
    "Context overflow (prompt truncated)",
    "Infrastructure", "Out-of-memory (Jetson OOM)",
    "Network / Redis error", "Database error",
    "User expectation gap", "Misunderstood intent",
    "Insufficient depth or detail"
  ),
  stringsAsFactors = FALSE
)

# Indent subcategories
taxonomy$Code <- ifelse(
  nchar(taxonomy$Code) > 1,
  paste0("  ", taxonomy$Code),
  taxonomy$Code
)

knitr::kable(
  taxonomy,
  format  = "latex",
  booktabs = TRUE,
  caption = "Failure taxonomy (6 top-level categories, 19 subcategories).",
  col.names = c("Code", "Description"),
  linesep = ""
) %>%
  kable_styling(font_size = 6, latex_options = c("hold_position", "scale_down")) %>%
  row_spec(c(1, 6, 11, 16, 19, 23), bold = TRUE)
```

# Results

## Overall Run Quality

```{r quality-bar, fig.cap="Run outcome distribution."}
if (n_total > 0) {
  quality_df <- runs %>%
    mutate(outcome = case_when(
      run_failed         ~ "Failed",
      is_negative        ~ "Negative feedback",
      is_positive        ~ "Positive feedback",
      TRUE               ~ "No feedback"
    )) %>%
    count(outcome) %>%
    mutate(pct = n / sum(n),
           outcome = fct_relevel(outcome, "Positive feedback",
                                 "No feedback", "Negative feedback", "Failed"))

  ggplot(quality_df, aes(x = outcome, y = pct, fill = outcome)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = paste0(n, "\n(", percent(pct, 1), ")")),
              vjust = -0.3, size = 2) +
    scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
    scale_fill_manual(values = c(
      "Positive feedback" = "#4caf50",
      "No feedback"       = "#9e9e9e",
      "Negative feedback" = "#ff9800",
      "Failed"            = "#f44336"
    )) +
    labs(x = NULL, y = "Proportion of runs", title = "Run Outcomes") +
    theme_paper() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 6))
} else {
  ggplot() + annotate("text", x=1, y=1, label="No data yet — collect runs first") +
    theme_void()
}
```

## Failure Category Frequency

```{r category-freq, fig.cap="Frequency of failure categories among coded runs."}
coded_runs <- runs %>% filter(!is.na(failure_category))

if (nrow(coded_runs) > 0) {
  cat_counts <- coded_runs %>%
    count(failure_cat_lbl) %>%
    filter(!is.na(failure_cat_lbl)) %>%
    mutate(pct = n / sum(n),
           failure_cat_lbl = fct_reorder(failure_cat_lbl, n))

  ggplot(cat_counts, aes(x = failure_cat_lbl, y = pct)) +
    geom_col(fill = "#1565c0", width = 0.6) +
    geom_text(aes(label = paste0(n, " (", percent(pct, 1), ")")),
              hjust = -0.1, size = 2) +
    coord_flip() +
    scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
    labs(x = NULL, y = "Share of coded failures",
         title = "Failure Category Distribution") +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No coded failures yet") +
    theme_void()
}
```

## Run Duration by Outcome

```{r duration-box, fig.cap="Run duration (s) by outcome category. Horizontal bars show medians."}
dur_df <- runs %>%
  filter(!is.na(duration_s)) %>%
  mutate(outcome = case_when(
    run_failed  ~ "Failed",
    is_negative ~ "Negative",
    is_positive ~ "Positive",
    TRUE        ~ "No feedback"
  ))

if (nrow(dur_df) > 0) {
  ggplot(dur_df, aes(x = outcome, y = duration_s, fill = outcome)) +
    geom_boxplot(outlier.size = 0.8, width = 0.5, show.legend = FALSE) +
    scale_fill_manual(values = c(
      "Positive"    = "#4caf50",
      "Negative"    = "#ff9800",
      "Failed"      = "#f44336",
      "No feedback" = "#9e9e9e"
    )) +
    labs(x = NULL, y = "Duration (seconds)", title = "Run Duration by Outcome") +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No duration data yet") +
    theme_void()
}
```

## LLM Call Count Distribution

```{r llm-calls-hist, fig.cap="Distribution of LLM calls per run by outcome."}
llm_df <- runs %>%
  filter(!is.na(llm_calls)) %>%
  mutate(outcome = case_when(
    run_failed  ~ "Failed",
    is_negative ~ "Negative",
    is_positive ~ "Positive",
    TRUE        ~ "No feedback"
  ))

if (nrow(llm_df) > 0) {
  ggplot(llm_df, aes(x = llm_calls, fill = outcome)) +
    geom_histogram(binwidth = 1, position = "stack", colour = "white", linewidth = 0.2) +
    scale_fill_manual(values = c(
      "Positive"    = "#4caf50",
      "Negative"    = "#ff9800",
      "Failed"      = "#f44336",
      "No feedback" = "#9e9e9e"
    )) +
    labs(x = "LLM calls per run", y = "Count",
         title = "LLM Call Distribution", fill = NULL) +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No LLM call data yet") +
    theme_void()
}
```

## Subcategory Breakdown

```{r subcat-table, results='asis'}
if (nrow(coded_runs) > 0) {
  subcat_tbl <- coded_runs %>%
    filter(!is.na(failure_subcategory)) %>%
    count(failure_category, failure_subcategory) %>%
    arrange(failure_category, desc(n)) %>%
    rename(Category = failure_category,
           Subcategory = failure_subcategory,
           Count = n)

  knitr::kable(
    subcat_tbl,
    format   = "latex",
    booktabs = TRUE,
    caption  = "Failure subcategory counts.",
    linesep  = ""
  ) %>%
    kable_styling(font_size = 6, latex_options = c("hold_position"))
} else {
  cat("*No coded failures yet — populate data/runs\\_coded.csv to see this table.*\n")
}
```

## Tool Usage and Failure Co-occurrence

```{r tool-failure, fig.cap="Tool invocation frequency for positive vs negative runs."}
tool_df <- runs %>%
  filter(!is.na(tools_used) & tools_used != "") %>%
  mutate(outcome = if_else(is_negative | run_failed, "Problematic", "OK")) %>%
  mutate(tool_list = str_split(tools_used, "; ")) %>%
  unnest(tool_list) %>%
  filter(!is.na(tool_list) & tool_list != "")

if (nrow(tool_df) > 0) {
  tool_counts <- tool_df %>%
    count(tool_list, outcome) %>%
    group_by(tool_list) %>%
    mutate(total = sum(n)) %>%
    ungroup() %>%
    mutate(tool_list = fct_reorder(tool_list, total))

  ggplot(tool_counts, aes(x = tool_list, y = n, fill = outcome)) +
    geom_col(position = "dodge", width = 0.6) +
    coord_flip() +
    scale_fill_manual(values = c("OK" = "#4caf50", "Problematic" = "#f44336")) +
    labs(x = NULL, y = "Invocation count",
         title = "Tool Usage by Run Outcome", fill = NULL) +
    theme_paper()
} else {
  ggplot() + annotate("text", x=1, y=1, label="No tool usage data yet") +
    theme_void()
}
```

# Discussion

```{r discussion-stats, include=FALSE}
# Pre-compute stats for inline use
pct_neg <- if (n_rated > 0) round(n_negative / n_rated * 100, 1) else NA
pct_fail <- if (n_total > 0) round(n_failed / n_total * 100, 1) else NA

top_cat <- coded_runs %>%
  count(failure_cat_lbl) %>%
  arrange(desc(n)) %>%
  slice(1) %>%
  pull(failure_cat_lbl)

top_cat <- if (length(top_cat) == 0) "N/A" else top_cat
```

Of the `r n_rated` rated runs, `r pct_neg`% received negative feedback and
`r pct_fail`% ended in technical failure. The most frequent failure category
was **`r top_cat`**. These results are consistent with prior reports that
sub-3B models struggle with structured output generation and multi-step
reasoning [@sub3b; @agentbench].

Several patterns warrant highlighting. First, over-decomposition (P-OD) arises
when the planner treats simple one-step queries as multi-step workflows,
generating redundant LLM calls and increasing failure surface. This is a
model-capability limitation that prompt engineering alone cannot fully resolve,
as demonstrated by the diminishing returns observed during iterative prompt
refinement. Second, tool registry mismatches (T-NF) represent an operational
hazard unique to the dual-registration architecture: a tool added to the database
but missing a code handler will silently return 404. A startup-time reconciliation
check would eliminate this class. Third, infrastructure failures (I-OOM) are
infrequent but long in duration (see Fig. 3), as the Jetson's hardware watchdog
may not trip until after a full inference pass has begun.

The in-situ feedback mechanism proved effective for naturalistic data collection:
the operator naturally left comments that described not just *what* failed but
*why*, providing richer signal than a binary label alone. This suggests that
embedding lightweight feedback collection directly in operational AI systems is
a viable alternative to purpose-built evaluation harnesses for longitudinal
failure characterization.

# Conclusion

We have presented a six-category, 19-subcategory taxonomy of failure modes for
a sub-3B agentic AI system on edge GPU hardware. Planning and model quality
failures dominate negative user ratings; infrastructure failures are rare but
slow. The taxonomy is derived from `r n_total` naturalistic runs and provides
a structured vocabulary for discussing, prioritizing, and mitigating agentic
system failures in resource-constrained deployments. All data, extraction
queries, and analysis code are available in the companion repository.

# References {-}
